---
title: "SWC Day 2: RStudio"
author: "Callie Chappell"
date: "3/29/2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Summary: 

#Import Data: 
```{r}
#install.packages('tidyverse')
#install.packages('rmarkdown')

#import gapminder: 
library(readr)
gapminder <- read_csv("data/gapminder-FiveYearData.csv")
head(gapminder)

```

#old way: 
```{r}
#Recap: subsetting 
x <- 1:10
x
x[2]
x[c(TRUE, FALSE)] #will return all odd numbers (will only give you values if you have true, which are the odds)

#Vectorization: 
x > 2 #will compare all values in vector to 2. 

#Vectorization and subsetting: 
x[x>3] #evaluate vector x, and return (subset) all values >3. 

#how does this work with a dataframe? 
head(gapminder)
#select rows in population < 1000, and get all of the columns 
gapminder[gapminder$pop <100000, ]
#select rows in population < 1000, and get some of the columns (2-4): 
gapminder[gapminder$pop <100000, c(2:4)]
```


##dplyr: 
Load dyplr: 
```{r}
#Dplyr: package inside of tidyverse 
library(tidyverse)
library(dplyr)
```

1. Select: select specific columns
```{r cars}
#old way: 
gapminder[,c('country', 'year')]
#dplyr way: dyplyr knows that country and year refers to columns and not variables (like gapminder)
select(gapminder, country, year)
```

2. Pipes: send data through multiple commands without creating intermediate outputs 
```{r}
#pipes: %>%
gapminder %>% select(country, year)
gapminder %>% select(ends_with('y'))
gapminder %>% select(-ends_with('y')) %>% head
```

3.Filter values by a specific value or quality
```{r}
#filter: 
#filter by year after 1990
gapminder %>% filter(year > 1990)
#filter after 1990 and by asia: 
gapminder %>% filter(year > 1990, continent == "Asia") %>% select(country, gdpPercap)

#can save the results as a variable: 
post1990 <- gapminder %>% filter(year > 1990)
head(post1990)
```
4. Group_by: group variables by a categorical variable 
```{r}
#Group-_by:
#groupby allows you to pick any variable you want to see: 
post1990 %>% group_by(continent)
```

5.Summarize: Create new variables or columns based on existing data
```{r}
#summarize: 
post1990 %>% summarize(mean_pop = mean(pop, na.rm = TRUE))
post1990 %>% group_by(continent) %>% summarize(country_count = n(), gdp_mean = mean(gdpPercap))
```

6. Mutate: Create new variables based on combinations of existing variables 
```{r}
gapminder %>% mutate(gdp = pop * gdpPercap)
gapminder %>% mutate(gdp = pop * gdpPercap) %>% mutate (gdp = gdp / 1000000)
```


