---
title: 'SWC Day 2: RStudio'
author: "Callie Chappell"
date: "3/29/2018"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Summary: 

#Import Data: 
```{r}
#install.packages('tidyverse')
#install.packages('rmarkdown')

#import gapminder: 
library(readr)
gapminder <- read_csv("data/gapminder-FiveYearData.csv")
head(gapminder)
```

#A quick recap on base R:  
Yesterday, we learned a lot about subsetting data using basic R syntax. Here's a quick reminder: 
```{r}
#Recap: subsetting 
x <- 1:10
x
x[2]
x[c(TRUE, FALSE)] #will return all odd numbers (will only give you values if you have true, which are the odds)

#Recap: vectorization: 
x > 2 #will compare all values in vector to 2. 

#Recap: combining vectorization and subsetting: 
x[x>3] #evaluate vector x, and return (subset) all values >3. 

#Recap: how does this work with a dataframe? 
head(gapminder)

#select rows in population < 1000, and get all of the columns 
gapminder[gapminder$pop <100000, ]

#select rows in population < 1000, and get some of the columns (2-4): 
gapminder[gapminder$pop <100000, c(2:4)]
```
But let's be honest...this is really confusing. A good (and MUCH more intuitive) alternative is dplyr! 

##dplyr: 
Load dyplr: 
```{r}
#Dplyr: package inside of tidyverse 
library(tidyverse)
library(dplyr)
```

dplyr is part of the tidyverse in R and is used to sort, filter, rearrange, and a bunch of other stuff. 

In this section, we'll be going over the following dyplyr verbs: 
- select
- pipes (%>%)
- filter
- group_by
- summarize
- mutate

P.S. for right now, you can ignore the "%>% head"" at the end of all of this stuff. All this does is just show the very top of the output so everything doesn't get all clogged up with data. Later, we'll learn that the %>% is called a pipe and actually is just the same as the | we learned about in the command line.   

Okay, let's go! 

1. Select: select specific columns
```{r cars}
#old way: 
gapminder[,c('country', 'year')]
#dplyr way: dyplyr knows that country and year refers to columns and not variables (like gapminder)
select(gapminder, country, year) %>% head
```

2. Pipes: send data through multiple commands without creating intermediate outputs 
```{r}
#pipes: %>%
gapminder %>% select(country, year) %>% head
gapminder %>% select(ends_with('y')) %>% head
gapminder %>% select(-ends_with('y')) %>% head
```

3.Filter values by a specific value or quality
```{r}
#filter: 
#filter by year after 1990
gapminder %>% filter(year > 1990) %>% head
#filter after 1990 and by asia: 
gapminder %>% filter(year > 1990, continent == "Asia") %>% select(country, gdpPercap) %>% head

#can save the results as a variable: 
post1990 <- gapminder %>% filter(year > 1990)
head(post1990)
```
4. Group_by: group variables by a categorical variable 
```{r}
#Group-_by:
#groupby allows you to pick any variable you want to see: 
post1990 %>% group_by(continent) %>% head
```

5.Summarize: Create new variables or columns based on existing data
```{r}
#summarize: 
post1990 %>% summarize(mean_pop = mean(pop, na.rm = TRUE)) %>% head
post1990 %>% group_by(continent) %>% summarize(country_count = n(), gdp_mean = mean(gdpPercap)) %>% head
```

6. Mutate: Create new variables based on combinations of existing variables 
```{r}
gapminder %>% mutate(gdp = pop * gdpPercap) %>% head
gapminder %>% mutate(gdp = pop * gdpPercap) %>% mutate (gdp = gdp / 1000000) %>% head
```

#ggplot: 
ggplot is a great way to create plots of your data! Luckily, it's part of the tidyverse, so yu already loaded it. 

1. Basic components of plotting in ggplot: 
  a. ggplot() = this calls the package ggplot (part of the tidyverse)
  b. ggplot(gapminder, ) = this tells ggplot what data you want to use 
  c. aes(x=, y=) = this is where you define what columns will be your x and y axes
  d. geom_point() = this is where you define what kind of plot you want: 
        geom_point() = scatter plot
        geom_bar() = bar plot 
        etc. 
      
Let's make our first plot: 
```{r}
library(ggplot2)
ggplot(gapminder, aes(y = lifeExp, x = gdpPercap)) +
  geom_point()
```
2. Now, we want to plot each country's life expectancy by year and by continent. 
Let's review a few things: 
1. We're using the group function in aes to group by country
2. We're using the color function in aes to color by continent 
3. We're using multiple geoms (geom_point and geom_line) to plot them on top of each other. Turns out, we can combine geoms. 

```{r}
ggplot(gapminder, aes(x = year, y = lifeExp, group = country, color = continent)) +
  geom_point() +
  geom_line()
```

3.Changing the color of points in ggplot:
A. Manually color points
```{r}
ggplot(gapminder, aes(y = lifeExp, x = gdpPercap)) +
  geom_point(color = "yellow")
```
B. Make points transparent: 
```{r}
ggplot(gapminder, aes(y = lifeExp, x = gdpPercap)) +
  geom_point(alpha = 0.2)
```

4. Scale data: 
```{r}
ggplot(gapminder, aes(y = lifeExp, x = gdpPercap)) +
  geom_point(alpha = 0.2) +
  scale_x_log10()
```

5. Title your graph and axess: 
```{r}
ggplot(gapminder, aes(x = year, y = lifeExp, group = country, color = continent)) +
  geom_point() +
  geom_line() +
  ggtitle("Time series of GDP per capita") + ylab("GDP per capita")
```

6. Saving your plot into a variable: 
```{r}
graph_life_gdp <- ggplot(gapminder, aes(y = lifeExp, x = gdpPercap)) +
  geom_point(alpha = 0.2) +
  scale_x_log10()
#it won't print your graphic (if saved as a variable) unless you call the variable
graph_life_gdp
```

7. Linear regression: use geom_smooth
Hey! You might have noticed that since we've saved the graph above as a variable (graph_life_gdp) above, we can now do a bunch of things to it (like adding the linear regression line). We could save this new graph with the linear regression line, if I wanted to. 
```{r}
graph_life_gdp + geom_smooth(method='loess') 
```
Now, let's put a second linear regression with the lm method (just in yellow) on top: 
```{r}
graph_life_gdp + geom_smooth(method='loess') + geom_smooth(method = 'lm', color = 'yellow')
```
Don't forget that you can go to the HELP tab and get help with the packages you're using (like how the heck do I get a confidence interval!?). 

8. Faceting: 
What if I want to compare multiple graphs in multiple conditions or subgroups. 
So let's take that graph_life_gdp:
```{r}
graph_life_gdp
```
This graphi s okay, but it's all different years and places all combined into one. 
What if I wanted to look at a bunch of different graphs but differentiated by year? (AND keep the axeses the same!!!) 
```{r}
graph_life_gdp + facet_wrap( ~ year)
```
THIS IS THE RAW, UNBRIDLED POWER OF FACET WRAP. 

We can also do the same fun stuff we did above, like add a linear regression line: 
```{r}
graph_life_gdp +
  facet_wrap( ~ year) +
  geom_smooth(method = 'lm')
```
9. Facet Grid: 
First, let's tie this back to dplyr: 

Make a new variable (using dyplyr) with the data filtered after 1997:
```{r}
post1997 <- gapminder %>% filter(year >=1997)
```
Take some code from above to re-plot (with some minor modifications). We're going to see facet_grid which makes new sub-graphs on 2 different categorical variables. Also, these work side-by-side (same axeses) which is INCREDIBLE!!! 
```{r}
ggplot(post1997, aes(y = lifeExp, x = gdpPercap)) +
  geom_point(alpha = 0.2) +
  facet_grid( year ~ continent)
```
#Assignment: Create a density plot of GDP per capita, filled by continent
```{r}
ggplot(gapminder, aes(gdpPercap, color = continent, fill = continent)) +
  geom_density(alpha = 0.2)
```
We can also add geom_run to give more granualar distribution:
```{r}
ggplot(gapminder, aes(gdpPercap, color = continent, fill = continent)) +
  geom_density(alpha = 0.2) +
  geom_rug (alpha = 0.5)
```
##tidyr: 

Import new "messed up" data file: 
```{r}
library(readr)
gapminder_wide <- read_csv("data/gapminder_wide.csv")
head(gapminder_wide)
```
Oh crap! This looks totally different from the first gapminder. How do we coax it into what we had before?!

First, let's make all of these gdpPercap_1952 columns into a new column with the old columns as rows as the new column. We're gonna separate out the name (using separate) into 
```{r}
gapminder_long <- gather(gapminder_wide, "key", "value", -country, - continent) %>% 
  separate(key, into = c('column_name', 'year'), sep = '_')
head(gapminder_long)
```
Now let's plot it (faceted by column volue, which are GDP per capita, life expectancy, and population) 
```{r}
ggplot(gapminder_long, aes(value, color = column_name)) +
  facet_wrap( ~ column_name, scales = "free") +
  geom_density()
```
Okay, last step: take all data that was in "value" column and put in a new column with whatever is in gdpPercap.
```{r}
gapminder_long %>% spread(column_name, value) %>% head()
```
Now we want to combine muliple things together: 
See, we can combine tidyr gather with ggplot! See, they can play nicely together. 
```{r}
gather(gapminder_wide, "key", "value", -country, - continent) %>% 
  separate(key, into = c('column_name', 'year'), sep = '_') %>%
  ggplot(aes(value)) +
  facet_wrap( ~ column_name, scales = "free") +
  geom_density()
```

##R Markdown: 
Okay, now we're going to get meta-meta. Talking about RMarkdown INSIDE of RMarkdown. Okay bear with me. If you've made it this far, you've probably already figured out how to make an RMarkdown (.rmd) file, but here goes. 

1. Making an RMarkdown file: 
a. Go to File --> New File --> RMarkdown. You might have to install RMarkdown. Do it! 
b. Enter the name of the document you want to make and your name. 
c. Go to town! 

2. Editing your rMarkdown file: 
You can insert your data and analysis in what is called "chunks". You can insert chunks of R code (plus other kinds of code, such as python, bash command line etc.). This can be found in the "insert" button at the top of your window. Just type your code in these chunks and write whatever you want (notes, etc.) outside of the chunks and you're good to go! 

